---
title: "acoustic_fts_models"
output: html_document
---

```{r}
library(tidyverse)
library(glmmTMB)

project_path <- "/Volumes/ECoG/k448_tempo_study"
fs <- 200  # Hz, used only for spike samples -> seconds

participants <- read_tsv(file.path(project_path, "rawdata", "participants.tsv"))
sessions <- read_tsv(file.path(project_path, "rawdata", "sessions.tsv")) %>% filter(exclude == 0)
meta <- left_join(sessions, participants, by = "participant_id")
meta$ID <- paste0(meta$participant_id, "_", meta$session)
```

# load spikes function
```{r}
load_spikes <- function(participant_id, session) {
  path <- file.path(project_path, "derivatives", "spike_detection",
                     participant_id, session,
                     paste0(participant_id, "_", session, "_spike_detection_all_spikes.xlsx"))
  spikes <- readxl::read_excel(path)
  spikes$spike_time_s <- spikes$spike_start / fs
  spikes
}
```

# load time markers function
```{r}
load_time_markers <- function(participant_id, session) {
  path <- file.path(project_path, "derivatives", "time_markers",
                     participant_id, session,
                     paste0(participant_id, "_", session, "_processed_time_markers.csv"))
  read_csv(path, show_col_types = FALSE)
}
```

# bin spikes into 1 second time bins
```{r}
bin_spikes_for_event <- function(spikes, event_row) {
  start_s <- floor(event_row$start_time)
  end_s   <- floor(event_row$end_time)
  bins <- tibble(second_in_stim = 0:(end_s - start_s - 1))
  bins$spike_count <- sapply(bins$second_in_stim, function(s) {
    b0 <- start_s + s
    sum(spikes$spike_time_s >= b0 & spikes$spike_time_s < b0 + 1)
  })
  bins
}
```

# load acoustic features and create offsets
```{r}
load_acoustic <- function(stimulus_name) {
  path <- file.path(project_path, "derivatives", "acoustic_ft_analysis",
                     paste0(stimulus_name, ".csv"))
  af <- read_csv(path, show_col_types = FALSE)
  af$second_in_stim <- 0:(nrow(af)-1)
  af
}
```

# Create full dataframe with participant spikes and acoustic features
```{r}
all_data <- list()

for (id in meta$ID) {
  participant_id <- meta$participant_id[meta$ID == id]
  session        <- meta$session[meta$ID == id]

  sp <- load_spikes(participant_id, session)
  tm <- load_time_markers(participant_id, session)

  stim_events <- tm %>%
  filter(!str_detect(event, "baseline|GAP|TALKING|SART")) %>%
  filter(!str_detect(event, regex("COLDPLAY|K448_MONO", ignore_case = TRUE)))

  for (i in seq_len(nrow(stim_events))) {
    event_row <- stim_events[i, ]
    stim_name <- event_row$event

    bins <- bin_spikes_for_event(sp, event_row)
    acoustic_fts <- load_acoustic(stim_name)
    merged <- left_join(bins, acoustic_fts, by = "second_in_stim") %>%
      filter(!is.na(RMS_1s_bins))   
    # filter drops trailing rows with no acoustic data - last two rows of acoustic features are NA 
    # 1) because they are only extracted from full seconds i.e. last partial second is not
    # and 2) modulation spectrum has a 2s windowing that prevents the penultimate full second to be analysed
    merged$ID <- id
    merged$piece <- stim_name
    all_data[[paste(id, stim_name)]] <- merged
  }
}

full_df <- bind_rows(all_data)
full_df %>% count(piece)
full_df %>% count(ID)
```

# Centre within piece and scale by each acoustic feature's own pooled SD
```{r}
full_df <- full_df %>%
  group_by(piece) %>%
  mutate(across(matches("_1s_bins$|^modband_"), ~ . - mean(., na.rm = TRUE), .names = "{.col}_c")) %>%
  ungroup() %>%
  mutate(across(matches("_1s_bins_c$|^modband_.*_c$"), ~ . / sd(., na.rm = TRUE), .names = "{.col}_z"))

full_df %>%
  count(near_floor = modband_38.5 < -35) %>%
  mutate(pct = n / sum(n) * 100)

hist(full_df$modband_38.5, breaks = 50)
```

# Continuous model (negative binomial)
```{r}
m_cont <- glmmTMB(
  spike_count ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
    modband_2.8_c_z + modband_38.5_c_z + 
    piece + 
    (1 | ID),
  data = full_df,
  family = nbinom2()
)
summary(m_cont)
```

# Convert spike counts to binary spike present yes/no
```{r}
full_df <- full_df %>%
  mutate(spike_present = as.integer(spike_count > 0))

# Check counts of 1s bins with n spikes (most bins have 0 spikes, 1/5 of bins have 1 spike etc.)
spike_freqs <- full_df %>% count(spike_count)
ggplot(data=spike_freqs, aes(x=spike_count, y=n)) + geom_col() + ggtitle("Frequency of specific spike counts in 1-second bins")
```

# Binary model
```{r}
m_binary <- glmmTMB(
  spike_present ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
    modband_2.8_c_z + modband_38.5_c_z + 
    piece + (1 | ID),
  data = full_df,
  family = binomial()
)
summary(m_binary)
```

```{r}
library(broom.mixed)

tidy_binary <- tidy(m_binary, effects = "fixed", conf.int = TRUE) %>%
  mutate(model = "Binary (spike present)")

tidy_cont <- tidy(m_cont, effects = "fixed", conf.int = TRUE) %>%
  mutate(model = "Count (NB)")

coefs <- bind_rows(tidy_binary, tidy_cont) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    or_irr = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high),
    term_clean = term %>%
      str_remove("_1s_bins") %>%
      str_remove("_c_z$"), 
    significant = conf.low > 1 | conf.high < 1
  )

ordered_terms <- c(
  "pieceWAGNER",
  "pieceK448_166BPM",
  "pieceK448_136BPM",
  "pieceK448_106BPM",
  "RMS",
  "Centroid",
  "Flux",
  "modband_2.8",
  "modband_38.5"
)

coefs <- coefs %>%
  mutate(term_clean = factor(term_clean, levels = ordered_terms))

ggplot(coefs, aes(x = or_irr, y = term_clean, color = significant, shape = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                   position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c(`TRUE` = "#D10056", `FALSE` = "grey70"),
                      guide = "none") +
  scale_shape_manual(values = c(16, 17)) +
  scale_x_log10(breaks = c(0.7, 0.8, 0.9, 1, 1.1, 1.2, 1.3)) +
  labs(x = "Odds Ratio / Incidence Rate Ratio (log scale)", y = NULL, shape = "Model",
       title = "Effect of acoustic features and piece on IED likelihood/count") +
  theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

```{r}
m_piece_bin <- glmmTMB(
  spike_present ~ piece + (1 | ID),
  data = full_df,
  family = binomial()
)
summary(m_piece_bin)


m_piece_cont <- glmmTMB(
  spike_count ~ piece + (1 | ID),
  data = full_df,
  family = nbinom2()
)
summary(m_piece_cont)
```





# Permutation model: testing effect of acoustic features on spikes
## Permutation model functions
```{r}
circular_shift <- function(x, shift) {
  # Rotates a vector by `shift` positions (values wrap from the end to the start)
  # This is to break the temporal alignment between an acoustic feature and spikes 
  # while preserving the feature's own values and internal autocorrelation.
  n <- length(x)
  shift <- shift %% n
  if (shift == 0) return(x)
  c(tail(x, n - shift), head(x, shift))
}

permute_once <- function(df) {
  # applies circular_shift to each feature column independently within every (participant, piece) group, 
  # using a different random shift per group. Produces one full permuted dataset where 
  # every piece's music-to-spike alignment has been scrambled, but each piece's own acoustic dynamics 
  # and each participant's own spike series are untouched.
  group_keys <- unique(df[c("ID", "piece")])
  pieces_shifted <- vector("list", nrow(group_keys))

  for (i in seq_len(nrow(group_keys))) {
    this_id <- group_keys$ID[i]
    this_piece <- group_keys$piece[i]

    sub_df <- df[df$ID == this_id & df$piece == this_piece, ]
    n <- nrow(sub_df)

    if (n < 10) {
      pieces_shifted[[i]] <- sub_df
      next
    }

    shift_amt <- sample(5:(n - 5), 1)
    for (term in feature_terms) {
      sub_df[[term]] <- circular_shift(sub_df[[term]], shift_amt)
    }

    pieces_shifted[[i]] <- sub_df
  }

  bind_rows(pieces_shifted)
}

get_zstats_both <- function(df) {
  # fits the binary (spike present/absent, logistic) and count (spike count, negative binomial) models
  # on the given dataset and returns the z-statistics for all feature terms from both models.
  m_binary <- glmmTMB(
    spike_present ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
      modband_2.8_c_z + modband_38.5_c_z + piece + (1 | ID),
    data = df, family = binomial()
  )
  m_count <- glmmTMB(
    spike_count ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
      modband_2.8_c_z + modband_38.5_c_z + piece + (1 | ID),
    data = df, family = nbinom2()
  )
  z_binary <- summary(m_binary)$coefficients$cond[feature_terms, "z value"]
  z_count  <- summary(m_count)$coefficients$cond[feature_terms, "z value"]
  names(z_binary) <- paste0("binary_", feature_terms)
  names(z_count)  <- paste0("count_", feature_terms)
  c(z_binary, z_count)
}

# returns the largest absolute z-statistic from a vector of z-statistics. 
# Applied to both the observed data and each permuted dataset to build the max-stat null distribution 
# used for multiple-comparisons correction.
max_stat <- function(z) max(abs(z), na.rm = TRUE)
```

## Testing a single permutation 
```{r}
feature_cols <- c("RMS_1s_bins", "Centroid_1s_bins", "Flux_1s_bins", "modband_2.8", "modband_38.5")
feature_terms <- paste0(feature_cols, "_c_z")

test_shift <- permute_once(full_df)
# should be very close to 0 and the means should be identical when sorted
mean(full_df$RMS_1s_bins_c_z, na.rm = TRUE); mean(test_shift$RMS_1s_bins_c_z, na.rm = TRUE)
# the two correlations should now be different
cor(full_df$RMS_1s_bins_c_z, full_df$spike_count, use = "complete.obs")
cor(test_shift$RMS_1s_bins_c_z, test_shift$spike_count, use = "complete.obs")
```

## Final permutation model
```{r}
n_perm <- 100
null_max <- numeric(n_perm)

set.seed(42)
for (p in seq_len(n_perm)) {
  shifted <- permute_once(full_df)
  z <- tryCatch(get_zstats_both(shifted), error = function(e) NA)
  null_max[p] <- if (all(!is.na(z))) max_stat(z) else NA
  if (p %% 10 == 0) message("permutation ", p, " done")
}

sum(is.na(null_max))  # check how many failed
```

## Actual model with observed z-scores
```{r}
observed_z <- get_zstats_both(full_df)
observed_max <- max_stat(observed_z)
```

## Check p values against the permuted model
```{r}
term_pvals <- sapply(abs(observed_z), function(oz) mean(null_max >= oz, na.rm = TRUE))

results <- tibble(
  term = names(observed_z),
  observed_z = observed_z,
  p_permutation = term_pvals
) %>% arrange(p_permutation)

results
```


# Permutation model: testing interactions between modspec 38.5 Hz and piece on spikes
```{r}
get_zstats_interaction <- function(df) {
  m_binary <- glmmTMB(
    spike_present ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
      modband_2.8_c_z + modband_38.5_c_z * piece + (1 | ID),
    data = df, family = binomial()
  )
  m_count <- glmmTMB(
    spike_count ~ RMS_1s_bins_c_z + Centroid_1s_bins_c_z + Flux_1s_bins_c_z +
      modband_2.8_c_z + modband_38.5_c_z * piece + (1 | ID),
    data = df, family = nbinom2()
  )

  interaction_terms <- grep("modband_38.5_c_z:piece",
                             rownames(summary(m_binary)$coefficients$cond),
                             value = TRUE)

  z_binary <- summary(m_binary)$coefficients$cond[interaction_terms, "z value"]
  z_count  <- summary(m_count)$coefficients$cond[interaction_terms, "z value"]
  names(z_binary) <- paste0("binary_", interaction_terms)
  names(z_count)  <- paste0("count_", interaction_terms)
  c(z_binary, z_count)
}
```

## Permuted model
```{r}
n_perm <- 100
null_max_interaction <- numeric(n_perm)

set.seed(43)
for (p in seq_len(n_perm)) {
  shifted <- permute_once(full_df)
  z <- tryCatch(get_zstats_interaction(shifted), error = function(e) NA)
  null_max_interaction[p] <- if (all(!is.na(z))) max_stat(z) else NA
  if (p %% 10 == 0) message("permutation ", p, " done")
}

sum(is.na(null_max_interaction))
```

## Real data model
```{r}
observed_z_interaction <- get_zstats_interaction(full_df)
observed_z_interaction
length(observed_z_interaction)  # should be 8
observed_max_interaction <- max_stat(observed_z_interaction)
observed_max_interaction
```

## Corrected p-values
```{r}
interaction_pvals <- sapply(abs(observed_z_interaction),
                             function(oz) mean(null_max_interaction >= oz, na.rm = TRUE))

interaction_results <- tibble(
  term = names(observed_z_interaction),
  observed_z = observed_z_interaction,
  p_permutation = interaction_pvals
) %>% arrange(p_permutation)

interaction_results
```

```{r}
full_df %>%
  group_by(piece) %>%
  summarise(n = n(), sd_38.5 = sd(modband_38.5, na.rm = TRUE), range_38.5 = diff(range(modband_38.5, na.rm = TRUE)))


full_df %>% filter(piece %in% c("WAGNER", "BACH")) %>%
  ggplot(aes(x = second_in_stim, y = modband_38.5, color = piece)) + geom_line() + facet_wrap(~piece, scales = "free_x")
```




